---
title: "Subsetting in Base R Tutorial"
author: "Dave Lovell"
date: "2023-12-12"
output: learnr::tutorial
runtime: shiny_prerendered
---


<!-- ## Libraries -->

```{r libraries, message = FALSE, include=FALSE}
library(learnr)
library(dplyr)
library(stringr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

x <- c(a = 1, b = 2, c = 3, d = 4)
```

# Subsetting operators

There are three subsetting operators in R: `[`, `[[` and `$`. `[[` and `$` are closely related to one another.

## Subsetting with `[`

I think of `[` as the most basic subsetting operator. You can use it in various ways:

### 1. With a numeric vector

```{r numeric_subset_basic}
x <- c(a = 1, b = 2, c = 3, d = 4)

x[4]
x[2:3]
x[c(1,3)]
```

### 2. With a logical vector

```{r logical_subset_basic}
x[c(TRUE, TRUE, TRUE, TRUE)]
x[c(TRUE, FALSE, FALSE, TRUE)]
```

Logical vectors are 'recycled' if they are shorter than the vector being subset:

```{r recycled}
x[c(TRUE, FALSE)]
x[c(TRUE, TRUE)]
x[TRUE]
x[FALSE]
```

### 3. With a character vector

If a vector has names, it can be subset with a character vector:

```{r character_subsetting}
x["a"]


```

