---
title: "Mutate Tutorial"
author: "Dave Lovell"
date: "2023-12-12"
output: learnr::tutorial
runtime: shiny_prerendered
---


<!-- ## Libraries -->

```{r libraries, message = FALSE, include=FALSE}
library(learnr)
library(dplyr)
library(stringr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

trivial <- 
  tibble(numbers = 1:4, letters = letters[1:4])
```


## Introduction

`mutate()` is one of the 'big 5' `dplyr` functions. These are:

* `select()` for picking columns
* `filter()` for picking rows
* `mutate()` for creating and modifying columns
* `summarise()` for summarising rows
* `reframe()` is just a variant of summarise really

### Making a trivial tibble 

Let's make a trivial tibble to practice mutating on:

```{r, trivial_tibble}
trivial <- 
  tibble(numbers = 1:4, letters = letters[1:4])

head(trivial)
```

## Required inputs & expected outputs

### Minimum function requirements:

Mutate only technically requires one argument, `.data`, which is the data frame on which the mutate operation will be performed. If no other arguments are specified, `mutate()` just returns `.data`.:

```{r most_basic_mutate_possible}
mutate(trivial) 
```

### Returned data type:

`mutate()` always returns an object with the same type as `.data`:

```{r mutating_a_tibble}
class(mutate(trivial))
```

```{r mutating_a_dataframe}
dataframe <- data.frame(a = 1:10)
class(mutate(dataframe))
```

### Accepted types for `.data`

`.data` must be a data frame or an extension thereof (e.g. a tibble):

```{r type_of_data}
foo <- matrix(1:4, nrow = 2)

tryCatch(mutate(foo), error = \(cnd) cnd$message)
```

### Summary

* `mutate()` needs a `.data` argument
* `.data` must be a type of dataframe
* `mutate()` always returns something with the same type as `.data`

## Actual mutate use

### Using the dots

Of course, `mutate()` most minimal behaviour is pretty useless, because it just returns whatever we put in. The real joy of mutate is in its powerful `...` argument. By passing *expressions* to  `...`, we can specify any number of new columns to add to our dataframe:

```{r using_dots}
mutate(trivial,
       new_numbers = numbers / 4,
       cool_numbers = new_numbers * 10)
```

### Evaluation environment

A lot of mutate's utility comes from the fact that anything passed to `...` is evaluated using the columns within `.data` as the primary evaluation environment. Hence these two calls are equivalent:

```{r evaluation_environment, eval = FALSE}
trivial$doubled_numbers <- trivial$numbers * 2

trivial <- mutate(trivial, doubled_numbers = numbers * 2)
```

With that in mind, can you write a `mutate()` operation that's equivalent to the following code in the tutorial block below?

```{r, eval = FALSE}
trivial$big_letters <- str_to_upper(trivial$letters)
```

```{r, mutate_equivalent, exercise = TRUE}
trivial <- mutate(trivial, )
```

```{r mutate_equivalent-solution}
trivial <- mutate(trivial, big_letters <- str_to_upper(letters))
```

### Variable lookup

If a variable is not found within the evaluation environment (i.e it's not the name of a column in `.data`), then R will search for it in the environment from which `mutate()` was called. 

Take a look at this code:

```{r environments_example, eval = FALSE}
letters <- c("z", "y", "x", "w")

mutate(trivial, big = str_to_upper(letters))
```

```{r what_will_big_b, echo = FALSE}
question(
  "What will `letters$big` be if the code is run?",
  answer("Z  Y  X  W"),
  answer("A  B  C  D", correct = TRUE),
  answer("NA"),
  answer("1  2  3  4")
)
```

## Ways to use `mutate()`

Mutate can be used to add new columns to a dataframe, or to replace existing columns. You can also operate on columns new columns within the same `mutate()` that creates them.

### Creating new columns

Add a new column to `trivial` that contains the values of `trivial$numbers` multiplied by three. Name it something sensible:

```{r doubling_numbers_new_column, exercise = TRUE}
mutate(trivial, )
```

### Overwriting existing columns

By specifying the column name as the name of an existing column, we can overwrite that column. Replace `trivial$numbers` with something else:

```{r replacing_numbers, exercise = TRUE}
mutate(trivial,  )
```

### Using/replacing columns within a single `mutate()`

Columns created as part of a `mutate()` call can used or replaced within the same mutate call:

```{r multiple_columns}
mutate(trivial,
       halved     = numbers / 2,
       doubled    = halved * 2 * 2,
       quadrupled = doubled * 2,
       halved = halved ^ 3
       )
```

## Grouped mutates

Sometimes we want to group certain rows together for a mutate operation. We can do this by using `dplyr::group_by()`. For example:

```{r group_by_demo}
trivial <- mutate(trivial, is_even = numbers %% 2 == 0)
trivial

trivial |> 
  group_by(is_even) |> 
  mutate(set_mean = mean(numbers))
```
The argument to `group_by()` is an expression, so `group_by(is_even)` could also be expressed as `group_by(is_even = numbers %% 2 == 0)`:

```{r explicit_demonstration_of_group_by_expression}
trivial |> 
  group_by(is_even = numbers %% 2 == 0) |> 
  mutate(set_mean = mean(numbers))
```

Can you group `trivial` by `numbers >= 3` and add a new column that contains the sum of each number with the `max()` of the group?

```{r group_by_test, exercise = TRUE}
trivial |> 
  group_by() |> 
  mutate(set_max_plus_num = )
```

## Any Questions?

## Other arguments to mutate

### `.by` (experimental)

Use this for 'inline' grouping. Note that this can only be a selection of columns, not an expression.

```{r by_eg}
trivial |> 
  mutate(is_even = numbers %% 2 == 0) |> 
  mutate(sum = sum(numbers), .by = is_even)
```

### .before and . after

New columns appear on the left-hand-side of the output by default. If you want a new column to appear somewhere else, you can use these:

```{r before_after_eg}
mutate(trivial, doubled = numbers * 2, .before = letters)
```

### .keep

Sometimes useful but it might confuse readers so bear that in mind. `.keep` Can be one of `c("all", "used", "unused", "none")` (`"all"` is the default):

```{r keep_eg}
## adding a grouping column to demonstrate how it's handled
trivial <- 
  mutate(trivial, is_even = numbers %% 2 == 0) |> 
  group_by(is_even)

mutate(trivial, mean = mean(numbers), .keep = "all") #default behaviour
mutate(trivial, mean = mean(numbers), .keep = "used")
mutate(trivial, mean = mean(numbers), .keep = "unused")
mutate(trivial, mean = mean(numbers), .keep = "none")
```

### Finally:

```{r finally}
mutate(trivial, sum = mean(numbers), .keep = "none") |> 
  distinct()

summarise(trivial, sum = mean(numbers))
```

## Quiz

```{r quiz, echo = FALSE}
quiz(
  caption = "End of session quiz",
  
  question(
    "Which of these is true of the object returned by `mutate()`",
    answer("It will always be a tibble"),
    answer("It will always have more columns that `.data`"),
    answer("It will always be a type of data frame", correct = TRUE),
    answer("It will always have at least one row")
  ),
  
  question(
    "How many named arguments does `mutate()` have?",
    answer("4"),
    answer("5"),
    answer("6", correct = TRUE),
    answer("7")
  ),
  
  question(
    "What's the first place that `mutate()` attempts to evaluate expressions in `...`?",
    answer("The global environment"),
    answer("An environment constructed from the columns of `data`", correct = TRUE)
  ),
  
  question(
    "If _n_ is the number of rows in `.data`, what are acceptable lenghts for expressions passed to `...`?",
    answer("Only n"),
    answer("Any length between 1 and n"),
    answer("Any length that is a factor of n"),
    answer("Any positive length is acceptable"),
    answer("One or n only", correct = TRUE)
  )
)